Polynomial Time and Dependent Types

### 2023-10-24 paper/paper.tex
[ paper ] typo
[ paper ] equational theory throughout

We will see in \autoref{sec:consfree-sound} that this system is sound
for polytime by a realisability argument. Completeness can be seen
more directly by constructing a function that iterates a function for
a statically known polynomial number of times in the size of the
input. Assume that we have a known polynomial
$p(n) = c_dn^d + \dots + c_0$ of degree $d$ with natural number
coefficients and some single step function
$f : \mathit{St} \multimap \mathit{St}$ over a state type
$\mathit{St}$ that runs to completion for input of size $n$ in $p(n)$
steps. Then, using the iterator above we can iterate $f$ over a
$\tyNat$ representing the size of the input:
\begin{displaymath}
  \begin{array}{l}
    I_1 : \tyNat \multimap \mathit{St} \multimap \mathit{St} \\
    I_1 = \lambda n. \lambda s.\tmRec\,n\,\{\conZero \mapsto s; \conSucc(s) \mapsto f\,s \}
  \end{array}
\end{displaymath}
To achieve higher degrees, we can use $\dupNat$ to nest iterations:
\begin{displaymath}
  \begin{array}{l}
    I_{k+1} : \tyNat \multimap \mathit{St} \multimap \mathit{St} \\
    I_{k+1} = \lambda n. \lambda s.
    \begin{array}[t]{@{}l}
      \Let\,(n,n') = \dupNat\,n\,\In\\
      \tmRec\,n\,\{\conZero \mapsto s; \conSucc(s) \mapsto I_k\,n'\,s \}
    \end{array}
  \end{array}
\end{displaymath}
By further use of $\dupNat$ and composition to handle addition of
polynomials, the function $f$ can now be iterated $p(n)$ many times,
where $n$ is the input $\tyNat$. Thus, the \emph{Cons-free} system can
represent all polytime computations.

@@

QTT is a system that resolves this conflict by recording usage of
variables with annotations from a semiring. It sits in the general
area of systems that use semiring annotatons to measure resource usage
\cite{BrunelGMZ14,GhicaS14,OrchardLE19}. The key feature of QTT, an
insight owing to \citet{mcbride16}, is that usage of variables in
types counts for $0$-usage in terms of the semiring used. This allows
us to use normal type theory as a specification language, while also
enjoying the benefits of linear typing for programs. The term typing
judgement of QTT has the following form:
\begin{displaymath}
  x_1 \stackrel{\rho_1}: S_1, \dots, x_n \stackrel{\rho_n}: S_n \vdash M \stackrel\sigma: T
\end{displaymath}
where the annotations $\rho_i$ are all from the semiring being
used. The annotation $\sigma$ is either $0$ or $1$, indicating whether
we are in the erased ($\sigma = 0$) fragment, where all the normal
rules of type theory apply, or the in the non-erased (``present'',
``realisable'', $\sigma = 1$) fragment, where a restricted typing
discipline applies.  As we shall see below in the cases of
$\Sigma$-types, iterable types, and LFPL's $\Diamond$ type, the
separation of QTT into these two fragments allows an expressive
combination of reasoning using full type theory with the benefits of
linear typing.

@@

\bob{Mention equality rules, and put them in an appendix; also
  Conversion rule is missing!}

@@

  \inferrule*
  {\Gamma_1 \vdash M \stackrel\sigma: (x \stackrel\rho: S) \to T \\
    \Gamma_2 \vdash N \stackrel{\sigma'}: S \\
    0\Gamma_1 = 0\Gamma_2 \\
    \sigma' = 0 \Leftrightarrow (\rho = 0 \lor \sigma = 0)}
  {\Gamma_1 + \rho\Gamma_2 \vdash M\,N \stackrel\sigma: T[N/x]}
\end{mathpar}
The side conditions on the elimination rule state that (i) both
$\Gamma_1$ and $\Gamma_2$ erase to the same context, so their sum is
defined; and (ii) the argument $N$ is erased (i.e., $\sigma' = 0$) iff
either the function does not use its argument, or we are in the
$\sigma = 0$ fragment and everything is being erased. In the
following, when we write $S \to T$ for a non-dependent function type,
we mean that the argument is to be used linearly:
$(x \stackrel1: S) \to T$, where $x$ does not appear in $T$.

\bob{$\beta\eta$-laws for $\Pi$-types}

@@

  \inferrule*
  {\Gamma \vdash M \stackrel0: (x \stackrel\pi: S) \otimes T}
  {\Gamma \vdash \mathrm{snd}(M) \stackrel0: T[\mathrm{fst}(M)/x]}
\end{mathpar}
$\Sigma$-types are unrestricted in the $\sigma = 0$ fragment, and we
can use them as normal for type-theoretic constructions.  In the
$\sigma = 1$ fragment, we must take into account the resource content
of objects and use a pattern matching construct; the dependently typed
analogue of the $\otimes$-eliminator in \autoref{sec:linear-intro}:
\begin{mathpar}
  \inferrule*
  {0\Gamma, z \stackrel0: (x \stackrel\pi: A) \otimes B \vdash C \\
    \Gamma_1 \vdash M \stackrel\sigma: (x \stackrel\pi: A) \otimes B \\
    \Gamma_2, x \stackrel{\sigma\pi}: A, y \stackrel{\sigma}: B \vdash N \stackrel\sigma: C[(x,y)/z]\\
    0\Gamma_1 = 0\Gamma_2}
  {\Gamma_1 + \Gamma_2 \vdash \mathrm{let}~(x,y) = M~\mathrm{in}~N \stackrel\sigma: C[M/z]}
\end{mathpar}
QTT also supports a unit type $I$ with constructor $*$ and
pattern-matching \cite{atkey18qtt}.

\bob{$\beta\eta$-laws for $\Sigma$-types, with wrinkle in the $\sigma = 1$ fragment}

@@

\subsubsection{Universe} QTT has universe types $\mathsf{U}$, as in
standard type theory \cite{atkey18qtt}.

@@

  \inferrule*
  {0\Gamma_1, x \stackrel0: \BoolTy \vdash P~\istype \\
    \Gamma_1 \vdash M \stackrel\sigma: \BoolTy \\
    \Gamma_2 \vdash N_t \stackrel\sigma: P[\cTrue/x] \\
    \Gamma_2 \vdash N_f \stackrel\sigma: P[\cFalse/x] \\
    0\Gamma_1 = 0\Gamma_2}
  {\Gamma_1 + \Gamma_2 \vdash \If_{x.P}\: M \: \Then \: N_t \: \Else \: N_f \stackrel\sigma: P[M/x]}
\end{mathpar}
\bob{Mention the $\beta$ law}
The introduction rules for booleans both use a $0$-d context,
indicating that construction of boolean values is free. Elimination of
booleans via a dependently typed if-then-else is more subtle with its
resource usage. The boolean to be eliminated must be constructed in a
context $\Gamma_1$, while the two branches are constructed in context
$\Gamma_2$. Since only one of the branches will be used, sharing
resources between the branches is expected.

@@

  \inferrule*
  {\Gamma_1 \vdash M \stackrel\sigma: T \\
    \Gamma_2 \vdash N \stackrel\sigma: \ListTy(T) \\
    0\Gamma_1 = 0\Gamma_2}
  {\Gamma_1 + \Gamma_2 \vdash \cCons(M,N) \stackrel\sigma: \ListTy(T)}
\end{mathpar}
Lists do have the potential for iteration by their recursive nature,
but in order to ensure the complexity guarantees we only permit
matching without recursion in the full calculus. Here is the rule for
dependently typed case analysis on lists:
\begin{mathpar}
  \inferrule*
  {0\Gamma_1, x \stackrel0: \ListTy(T) \vdash P~\istype \\
    \Gamma_1 \vdash M \stackrel\sigma: \ListTy(T) \\
    \Gamma_2 \vdash N_1 \stackrel\sigma: P[\cNil/x] \\
    \Gamma_2, h \stackrel\sigma: T, t \stackrel\sigma: \ListTy(T) \vdash N_2 \stackrel\sigma: P[\cCons(h,t)/x] \\
    0\Gamma_1 = 0\Gamma_2}
  {\Gamma_1 + \Gamma_2 \vdash \Match_{x.P}\,M\,\{\,\cNil \mapsto N_1; \cCons(h,t) \mapsto N_2\,\} \stackrel\sigma: P[M/x]}
\end{mathpar}
\bob{$\beta$ laws!}
In the $\sigma = 0$ fragment, however, we are free to iterate on lists
because computations in this fragment are only meant for type-level
computation, not for the program itself. Put another way, the type
checker may perform arbitary recursion on lists to type check the
program, but the program itself may not do so without correctly
accounting its costs as described in the following sections. The
$\sigma = 0$ recursor for lists has the following typing rule, which
is the standard dependent eliminator for lists, with everything
annotated as $0$ usage.
\begin{mathpar}
  \inferrule*
  {0\Gamma, x \stackrel0: \ListTy(T) \vdash P~\istype \\
    0\Gamma \vdash M \stackrel0: \ListTy(T) \\
    0\Gamma \vdash N_1 \stackrel0: P[\cNil/x] \\
    0\Gamma, h \stackrel0: T, t \stackrel0: \ListTy(T), p \stackrel0: P[t/x] \vdash N_2 \stackrel0: P[\cCons(h,t)/x]}
  {0\Gamma \vdash \mathrm{recList}_{x.P}\,M\,\{\,\cNil \mapsto N_1; \cCons(h,t;p) \mapsto N_2\,\} \stackrel0: P[M/x]}
\end{mathpar}

@@

The datatypes of the previous section still only allow us to write
programs in the $\sigma = 1$ fragment that are constant time in the
size of their input. As with the simply typed linear system, if we are
handed a list of an unknown length, we can only explore it to a fixed
depth determined statically. To write programs that do work
proportional to the size of their input, we need some form of iterable
datatype. In both our Cons-free and LFPL-style QTT systems, we use a
natural number datatype.

@@

  \inferrule*
  {\Gamma \vdash M \stackrel0: \tyNat}
  {\Gamma \vdash \conSucc(M) \stackrel0: \tyNat}
\end{mathpar}
The cons-free system allows free duplication of complete natural
numbers. This is accomplished by a special construct copying the
simply linear typed rule we gave above:
\begin{mathpar}
  \inferrule*
  {\Gamma \vdash M \stackrel\sigma: \tyNat}
  {\Gamma \vdash \mathrm{dupNat}(M) \stackrel\sigma: \tyNat \otimes \tyNat}
\end{mathpar}
Anyone who has reasoned about the metatheory of or implemented a type
checker for dependent types will view this rule with unease as it
appears to grant the ability to construct non-canonical values of pair
type, and consequently generate non-canonical naturals. We fix this by
adding an equational rule to the $\sigma = 0$ fragment, ensuring
\emph{definitionally} that $\mathrm{dupNat}$ acts as its name implies:
\begin{mathpar}
  \inferrule*
  {\Gamma \vdash M \stackrel0: \tyNat}
  {\Gamma \vdash \mathrm{dupNat}(M) \equiv (M,M) \stackrel0: \tyNat \otimes \tyNat}
\end{mathpar}
Note that this rule is well-typed by the $0$-needs-$0$ property of
QTT, and the fact that $0+0=0$.

The eliminator for these natural numbers takes the following
form. Disregarding the usage annotations, this has the same type
structure as the normal dependently typed recursor for natural
numbers:
\begin{displaymath}
  \mprset{flushleft}
  \inferrule*
  {0\Gamma, x \stackrel0: \tyNat \vdash P~\istype \\\\
    \Gamma \vdash M \stackrel\sigma: \tyNat \\\\
    0\Gamma \vdash N_z \stackrel\sigma: P[\conZero/x] \\\\
    0\Gamma, n \stackrel0: \tyNat, p \stackrel\sigma: P[n/x] \vdash N_s \stackrel\sigma: P[\conSucc(n)/x]}
  {\Gamma \vdash \tmRec_{x.P}\,M\,\{\conZero \mapsto N_z; \conSucc(n;p) \mapsto N_s\} \stackrel\sigma: P[M/x]}
\end{displaymath}
In the successor case, $N_s$, there are two bound variables: $n$ for
the natural number and $p$ for its induction hypothesis. Note that $n$
is required to be usage $0$ no matter what $\sigma$ is. We need the
variable $n$ to be present in order to correctly type the induction
hypothesis and the conclusion, but it must be marked as usage $0$ to
ensure that the resources captured by the number are not duplicated.

\bob{$\beta$ laws in the $\sigma = 0$ fragment}

@@

\bob{$\beta$ laws in both fragments}

@@

\bob{Equality rules (see response)}

>>>>

We will see in \autoref{sec:consfree-sound} that this system is sound
for polytime by a realisability argument. Completeness can be seen
more directly by constructing a function that iterates a function for
a statically known polynomial number of times in the size of the
input. Assume that we have a known polynomial
$p(n) = c_dn^d + \dots + c_0$ of degree $d$ with natural number
coefficients and some single step function
$f : \mathit{St} \multimap \mathit{St}$ over a state type
$\mathit{St}$ that runs to completion for input of size $n$ in $p(n)$
steps. Then, using the iterator above we can iterate $f$ over a
$\tyNat$ representing the size of the input:\bob{CHECK THIS}
\begin{displaymath}
  \begin{array}{l}
    I_1 : \tyNat \multimap \mathit{St} \multimap \mathit{St} \\
    I_1 = \lambda n. \lambda s.\tmRec\,n\,\{\conZero \mapsto s; \conSucc(s) \mapsto f\,s \}
  \end{array}
\end{displaymath}
To achieve higher degrees, we can use $\dupNat$ to nest iterations:
\begin{displaymath}
  \begin{array}{l}
    I_{k+1} : \tyNat \multimap \mathit{St} \multimap \mathit{St} \\
    I_{k+1} = \lambda n. \lambda s.
    \begin{array}[t]{@{}l}
      \Let\,(n,n') = \dupNat\,n\,\In\\
      \tmRec\,n\,\{\conZero \mapsto s; \conSucc(s) \mapsto I_k\,n'\,s \}
    \end{array}
  \end{array}
\end{displaymath}
By further use of $\dupNat$ and composition to handle addition of
polynomials, the function $f$ can now be iterated $p(n)$ many times,
where $n$ is the input $\tyNat$. Thus, the \emph{Cons-free} system can
represent all polytime computations.

@@

QTT is a system that resolves this conflict by recording usage of
variables with annotations from a semiring. It sits in the general
area of systems that use semiring annotations to measure resource usage
\cite{BrunelGMZ14,GhicaS14,OrchardLE19}. The key feature of QTT, an
insight owing to \citet{mcbride16}, is that usage of variables in
types counts for $0$-usage in terms of the semiring used. This allows
us to use normal type theory as a specification language, while also
enjoying the benefits of linear typing for programs. The term typing
judgement of QTT has the following form:
\begin{displaymath}
  x_1 \stackrel{\rho_1}: S_1, \dots, x_n \stackrel{\rho_n}: S_n \vdash M \stackrel\sigma: T
\end{displaymath}
where the annotations $\rho_i$ are all from the semiring being
used. The annotation $\sigma$ is either $0$ or $1$, indicating whether
we are in the erased ($\sigma = 0$) fragment, where all the normal
rules of type theory apply, or the in the non-erased (``present'',
``realisable'', $\sigma = 1$) fragment, where a restricted typing
discipline applies.  As we shall see below in the cases of
$\Sigma$-types, iterable types, and LFPL's $\Diamond$ type, the
separation of QTT into these two fragments allows an expressive
combination of reasoning using full type theory with the benefits of
linear typing.

@@

As usual, definitional equality of types impacts typing of terms via
the conversion rule:
\begin{displaymath}
  \inferrule*
  {\Gamma \vdash M : S \\ 0\Gamma \vdash S \equiv T~\istype}
  {\Gamma \vdash M : T}
\end{displaymath}
Like type formation, definitional equality of types always takes place
in $0$d contexts. We will describe the definitional equality rules for
terms of each type as we introduce them. In QTT, it is possible for
the definitional equality of terms to differ between the $\sigma = 0$
and $\sigma = 1$ fragments, as we will see below.

@@

  \inferrule*
  {\Gamma_1 \vdash M \stackrel\sigma: (x \stackrel\rho: S) \to T \\
    \Gamma_2 \vdash N \stackrel{\sigma'}: S \\
    0\Gamma_1 = 0\Gamma_2 \\
    \sigma' = 0 \Leftrightarrow (\rho = 0 \lor \sigma = 0)}
  {\Gamma_1 + \rho\Gamma_2 \vdash M\,N \stackrel\sigma: T[N/x]}
\end{mathpar}
The side conditions on the elimination rule state that (i) both
$\Gamma_1$ and $\Gamma_2$ erase to the same context, so their sum is
defined; and (ii) the argument $N$ is erased (i.e., $\sigma' = 0$) iff
either the function does not use its argument, or we are in the
$\sigma = 0$ fragment and everything is being erased. In the
following, when we write $S \to T$ for a non-dependent function type,
we mean that the argument is to be used linearly:
$(x \stackrel1: S) \to T$, where $x$ does not appear in
$T$. $\Pi$-types support the usual $\beta\eta$ definitional
equalities.

@@

  \inferrule*
  {\Gamma \vdash M \stackrel0: (x \stackrel\pi: S) \otimes T}
  {\Gamma \vdash \mathrm{snd}(M) \stackrel0: T[\mathrm{fst}(M)/x]}
\end{mathpar}
$\Sigma$-types are unrestricted in the $\sigma = 0$ fragment, and we
can use them as normal for type-theoretic constructions.  In the
$\sigma = 1$ fragment, we must take into account the resource content
of objects and use a pattern matching construct; the dependently typed
analogue of the $\otimes$-eliminator in \autoref{sec:linear-intro}:
\begin{mathpar}
  \inferrule*
  {0\Gamma, z \stackrel0: (x \stackrel\pi: A) \otimes B \vdash C \\
    \Gamma_1 \vdash M \stackrel\sigma: (x \stackrel\pi: A) \otimes B \\
    \Gamma_2, x \stackrel{\sigma\pi}: A, y \stackrel{\sigma}: B \vdash N \stackrel\sigma: C[(x,y)/z]\\
    0\Gamma_1 = 0\Gamma_2}
  {\Gamma_1 + \Gamma_2 \vdash \mathrm{let}~(x,y) = M~\mathrm{in}~N \stackrel\sigma: C[M/z]}
\end{mathpar}
QTT also supports a unit type $I$ with constructor $*$ and
pattern-matching \cite{atkey18qtt}. $\Sigma$- and $I$ types support
the usual $\beta\eta$ definitional equalities in the $\sigma = 0$
fragment (e.g., $\mathrm{fst}(M,N) \equiv M$), but only $\beta$
equalities (i.e.,
$\mathrm{let}~(x,y) = (M_1,M_2)~\mathrm{in}~N \equiv N[M_1/x,M_2/y]$)
in the $\sigma = 1$ fragment. It would also be sound to support
commuting conversions \cite{Barber1996} for $\mathrm{let}$ in the
$\sigma = 1$ fragment, but this would likely bring complications for
implementation.

@@

\subsubsection{Universe} QTT has universe types $\mathsf{U}$, as in
standard type theory \cite{atkey18qtt}. For our examples below, we do
not explicitly mark the use of terms of type $\mathsf{U}$ as types --
i.e., we use a Russell-style presentation. Universes are where the
definitional equality on terms affects the definitional equality on
types.

@@

  \inferrule*
  {0\Gamma_1, x \stackrel0: \BoolTy \vdash P~\istype \\
    \Gamma_1 \vdash M \stackrel\sigma: \BoolTy \\
    \Gamma_2 \vdash N_t \stackrel\sigma: P[\cTrue/x] \\
    \Gamma_2 \vdash N_f \stackrel\sigma: P[\cFalse/x] \\
    0\Gamma_1 = 0\Gamma_2}
  {\Gamma_1 + \Gamma_2 \vdash \If_{x.P}\: M \: \Then \: N_t \: \Else \: N_f \stackrel\sigma: P[M/x]}
\end{mathpar}
The introduction rules for booleans both use a $0$-d context,
indicating that construction of boolean values is free. Elimination of
booleans via a dependently typed if-then-else is more subtle with its
resource usage. The boolean to be eliminated must be constructed in a
context $\Gamma_1$, while the two branches are constructed in context
$\Gamma_2$. Since only one of the branches will be used, sharing
resources between the branches is expected. Booleans and their
eliminator obey the usual $\beta$ laws for definitional equality:
$\If_{x.P}\: \cTrue \: \Then \: N_t \: \Else \: N_f \equiv N_t$, and
similarly for $\cFalse$.

@@

  \inferrule*
  {\Gamma_1 \vdash M \stackrel\sigma: T \\
    \Gamma_2 \vdash N \stackrel\sigma: \ListTy(T) \\
    0\Gamma_1 = 0\Gamma_2}
  {\Gamma_1 + \Gamma_2 \vdash \cCons(M,N) \stackrel\sigma: \ListTy(T)}
\end{mathpar}
Lists do have the potential for iteration by their recursive nature,
but in order to ensure the complexity guarantees we only permit
matching without recursion in the full calculus. Here is the rule for
dependently typed case analysis on lists, which also obeys the usual
$\beta$-equalities for case analysis, analogous to the ones for
Booleans:
\begin{mathpar}
  \inferrule*
  {0\Gamma_1, x \stackrel0: \ListTy(T) \vdash P~\istype \\
    \Gamma_1 \vdash M \stackrel\sigma: \ListTy(T) \\
    \Gamma_2 \vdash N_1 \stackrel\sigma: P[\cNil/x] \\
    \Gamma_2, h \stackrel\sigma: T, t \stackrel\sigma: \ListTy(T) \vdash N_2 \stackrel\sigma: P[\cCons(h,t)/x] \\
    0\Gamma_1 = 0\Gamma_2}
  {\Gamma_1 + \Gamma_2 \vdash \Match_{x.P}\,M\,\{\,\cNil \mapsto N_1; \cCons(h,t) \mapsto N_2\,\} \stackrel\sigma: P[M/x]}
\end{mathpar}
In the $\sigma = 0$ fragment, however, we are free to iterate on lists
because computations in this fragment are only meant for type-level
computation, not for the program itself. Put another way, the type
checker may perform arbitary recursion on lists to type check the
program, but the program itself may not do so without correctly
accounting its costs as described in the following sections. The
$\sigma = 0$ recursor for lists has the following typing rule, which
is the standard dependent eliminator for lists, with everything
annotated as $0$ usage.
\begin{mathpar}
  \inferrule*
  {0\Gamma, x \stackrel0: \ListTy(T) \vdash P~\istype \\
    0\Gamma \vdash M \stackrel0: \ListTy(T) \\
    0\Gamma \vdash N_1 \stackrel0: P[\cNil/x] \\
    0\Gamma, h \stackrel0: T, t \stackrel0: \ListTy(T), p \stackrel0: P[t/x] \vdash N_2 \stackrel0: P[\cCons(h,t)/x]}
  {0\Gamma \vdash \mathrm{recList}_{x.P}\,M\,\{\,\cNil \mapsto N_1; \cCons(h,t;p) \mapsto N_2\,\} \stackrel0: P[M/x]}
\end{mathpar}
This eliminator also obeys the usual $\beta$-equality laws for a list
eliminator, using the resource freedom of the $\sigma = 0$ fragment to
duplicate the $N_2$ term in the $\cCons$ case.

@@

The datatypes of the previous section still only allow us to write
programs in the $\sigma = 1$ fragment that are constant time in the
size of their input. As with the simply typed linear system, if we are
handed a list of an unknown length, we can only explore it to a fixed
depth, determined statically by the program. To write programs that do
work proportional to the size of their input, we need some form of
iterable datatype. In both our Cons-free and LFPL-style QTT systems,
we use a natural number datatype.

@@

  \inferrule*
  {\Gamma \vdash M \stackrel0: \tyNat}
  {\Gamma \vdash \conSucc(M) \stackrel0: \tyNat}
\end{mathpar}
The cons-free system allows free duplication of complete natural
numbers. This is accomplished by a special construct copying the
simply linear typed rule we gave above:
\begin{mathpar}
  \inferrule*
  {\Gamma \vdash M \stackrel\sigma: \tyNat}
  {\Gamma \vdash \mathrm{dupNat}(M) \stackrel\sigma: \tyNat \otimes \tyNat}
\end{mathpar}
Anyone who has reasoned about the metatheory of, or implemented a type
checker for, dependent types will view this rule with unease as it
appears to grant the ability to construct non-canonical values of pair
type, and consequently generate non-canonical naturals. We fix this by
adding an equational rule to the $\sigma = 0$ fragment, ensuring
\emph{definitionally} that $\mathrm{dupNat}$ acts as its name implies:
\begin{mathpar}
  \inferrule*
  {\Gamma \vdash M \stackrel0: \tyNat}
  {\Gamma \vdash \mathrm{dupNat}(M) \equiv (M,M) \stackrel0: \tyNat \otimes \tyNat}
\end{mathpar}
Note that this rule is well-typed by the $0$-needs-$0$ property of
QTT, and the fact that $0+0=0$.

The eliminator for these natural numbers takes the following
form. Disregarding the usage annotations, it has the same type
structure as the normal dependently typed recursor for naturals:
\begin{displaymath}
  \mprset{flushleft}
  \inferrule*
  {0\Gamma, x \stackrel0: \tyNat \vdash P~\istype \\\\
    \Gamma \vdash M \stackrel\sigma: \tyNat \\\\
    0\Gamma \vdash N_z \stackrel\sigma: P[\conZero/x] \\\\
    0\Gamma, n \stackrel0: \tyNat, p \stackrel\sigma: P[n/x] \vdash N_s \stackrel\sigma: P[\conSucc(n)/x]}
  {\Gamma \vdash \tmRec_{x.P}\,M\,\{\conZero \mapsto N_z; \conSucc(n;p) \mapsto N_s\} \stackrel\sigma: P[M/x]}
\end{displaymath}
In the successor case, $N_s$, there are two bound variables: $n$ for
the natural number and $p$ for its induction hypothesis. Note that $n$
is required to be usage $0$ no matter what $\sigma$ is. We need the
variable $n$ to be present in order to correctly type the induction
hypothesis and the conclusion, but it must be marked as usage $0$ to
ensure that the resources captured by the number are not duplicated.

This eliminator cannot have a $\beta$-equality in the $\sigma = 1$
fragment because there is no way to construct any natural numbers to
iterate on in this fragment. In the $\sigma = 0$ fragment, this
eliminator obeys the usual $\beta$-equality laws for a natural number
recursor. This allows us to use it to compute and reason about
operations on naturals in this fragment.

@@

Unlike the natural number iterator in the Cons-free system, this
iterator has $\beta$-equalities in both the $\sigma = 0$ and
$\sigma = 1$ fragments. In the $\conSucc$ case, for example, we have:
\begin{displaymath}
  \begin{array}{cl}
    & \tmRec\,(\conSucc(M_d,M_n)\,\{\conZero(d) \mapsto N_z; \conSucc(d,n;p) \mapsto N_s\} \\
    \equiv & N_s[M_d/d, M_n/n, \tmRec\,M_n\,\{\conZero(d) \mapsto N_z; \conSucc(d,n;p) \mapsto N_s\}/p]
  \end{array}
\end{displaymath}
Note that the fact that the variable $n$ in the $N_s$ term is
annotated $0$, which allows us to duplicate it even when we are in the
$\sigma = 1$ fragment.

@@

The equality rules for $\Rtype$ state that the two operations are
mutually inverse: $\rIntro(\rElim(M)) \equiv M$, in both fragments,
and $\rElim(\rIntro(M)) \equiv M$ in the $\sigma = 0$ fragment.
